---
title: "pbmc68k.rmd"
output:
  html_document: default
  pdf_document: default
date: "2025-04-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Downloading data 68k PBMC

This notebook provides analysis pipeline to the dataset published in Zheng et al. 2017, called 'Fresh 68k PBMCs (Donor A)'.

Download the following data on 10X Genomics <https://support.10xgenomics.com/single-cell-gene-expression/datasets>.

The goal of Zheng et al. analysis was to profile 68k fresh peripheral blood mononuclear cells (PBMCs) from a healthy donor and demonstrated the scRNA-seq platform’s ability to dissect large immune populations.

```{r init}
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2) 

# Load the PBMC 68k dataset
pbmc.data <- Read10X(data.dir = "fresh_68k_pbmc_donor_a_filtered_gene_bc_matrices/filtered_matrices_mex/hg19/")

# Initialize the Seurat object with the raw (non-normalized data)
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc68k", min.cells = 3, min.features = 200)
pbmc
```

## Preprocessing data

```{r mito, fig.height=7, fig.width=13}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
```

```{r qc, fig.height=7, fig.width=13}

# Visualize QC metrics as a violin plot
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

# FeatureScatter is typically used to visualize feature-feature relationships, but can be used for anything calculated by the object, i.e. columns in object metadata, PC scores etc.

plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt") 
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") 
plot1 + plot2

pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)

# Visualize distribution of number of genes and UMI counts detected per 68k PBMCs after filtering
vplot <- VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA"), pt.size = 0, ncol = 2 , combine = FALSE, cols = c("#FFD1DC", "#FFD1DC"))

vplot[[1]] <- vplot[[1]] +  xlab("Genes/Cell") + ylab("Counts") + ggtitle(NULL) + 
  theme(
    axis.title.y = element_blank(),
    legend.position = "none",
    plot.title = element_blank()
  )

vplot[[2]] <- vplot[[2]] + xlab("UMIs Counts/Cell") + ggtitle(NULL) + 
  theme(
    axis.title.y = element_blank(),
    legend.position = "none",
    plot.title = element_blank()
  )


vln_plot <- wrap_plots(vplot, ncol = 2) + plot_annotation(theme = theme(plot.margin = margin(10, 10, 10, 10)))

vln_plot # save it to final figure
```

```{r normalize.default, eval = FALSE}
pbmc <- NormalizeData(pbmc)
```

# Identification of highly variable features (feature selection)

```{r hvf, fig.height=5, fig.width=11}
pbmc <- FindVariableFeatures(pbmc, selection.method = 'vst', nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(pbmc), 10)

# plot variable features with and without labels
plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```

## Scaling the data

```{r regress, fig.height=7, fig.width=11, results='hide'}
pbmc <- ScaleData(pbmc, features = VariableFeatures(pbmc))
```

## Perform linear dimensional reduction

```{r pca,results='hide'}
top_var_genes <- head(VariableFeatures(pbmc), 1000)
pbmc <- RunPCA(pbmc, features = top_var_genes)
```

```{r pca_viz, message=TRUE}
print(pbmc[['pca']], dims = 1:5, nfeatures = 5)
VizDimLoadings(pbmc, dims = 1:2, reduction = 'pca')
DimPlot(pbmc, reduction = 'pca')
```

## Determine the 'dimensionality' of the dataset

```{r}
ElbowPlot(pbmc, ndims = 30)
```

## Run non-linear dimensional reduction tSNE

```{r tsne, fig.height=5, fig.width=7}
pbmc <- RunTSNE(pbmc, dims = 1:10)

DimPlot(pbmc, reduction = 'tsne', label = TRUE, pt.size = 0.5)
```

```{r tsneplot_1, fig.height=5, fig.width=7}
tsne_coords <- Embeddings(pbmc, "tsne")
km.res <- kmeans(tsne_coords, centers = 10)
pbmc@meta.data$kmeans_clusters <- as.factor(km.res$cluster)  

# Calculate cluster percentages
cluster_percents <- prop.table(table(pbmc$kmeans_clusters)) * 100
percent_labels <- paste0(names(cluster_percents), " (", round(cluster_percents, 1), "%)")
cluster_labels <- setNames(percent_labels, names(cluster_percents))

# Plot with percentage labels
tsne_plot <- DimPlot(pbmc, 
                    reduction = "tsne", 
                    group.by = "kmeans_clusters",
                    label = TRUE,
                    pt.size = 0.5) +
  scale_color_discrete(labels = percent_labels) +
  theme(legend.position = "right",
        legend.title = element_blank(),
        plot.title = element_blank()) +
  guides(color = guide_legend(override.aes = list(size = 3)))

tsne_plot

```

## Finding differentially expressed features (cluster biomarkers)

```{r top2, fig.height=5, fig.width=7}
# Find top 2 markers for every cluster compared to all remaining cells, report only the positive ones
Idents(pbmc) <- "kmeans_clusters"
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE, min.pct = 0.25,  logfc.threshold = 0.25)
top2 <- pbmc.markers %>% group_by(cluster) %>% top_n(n = 2, wt = avg_log2FC)
```

```{r heatmap, fig.height=5, fig.width=7}
# Generate heatmap of top 2 markers per cluster
heatmap <- DoHeatmap(pbmc,
                     features = top2$gene,
                     group.by = "kmeans_clusters") +
  scale_fill_gradient2(low = "slategray2", mid = "slateblue1", high = "slateblue4", midpoint = 0) + theme(legend.title = element_blank()) +
  guides(color = guide_legend(FALSE))

heatmap
```

## Reproducing Figure 3

Reproducing Figure 3 from publication <https://www.nature.com/articles/ncomms14049/figures/3>.

Figure contains:

1.  Distribution of number of genes (and UMI counts detected per 68k PBMCs (vln_plot object).
2.  tSNE projection of 68k PBMCs, where each cell is grouped into one of the 10 clusters (distinguished by their colours) (tsne_plot object).
3.  Normalized expression (centred) of the top variable genes (rows) from each of 10 clusters (columns) is shown in a heatmap (heatmap object).
4.  tSNE projection of 68k PBMCs, with each cell coloured based on their normalized expression of *CD3D*, *CD8A*, *NKG7*, *FCER1A*, *CD16* and *S100A8* (are generated below).

```{r exp_proj, fig.height=5, fig.width=7}
# Generate tSNE projection of 68k PBMCs, with each cell coloured based on their normalized expression of CD3D, CD8A, NKG7, FCER1A, CD16 and S100A8. 

d <- FeaturePlot(pbmc, features = "CD3D", cols = c("lightblue", "lightpink")) + ggtitle("CD3D")
e <- FeaturePlot(pbmc, features = "CD8A", cols = c("lightblue", "lightpink")) + ggtitle("CD8A")
f <- FeaturePlot(pbmc, features = "NKG7", cols = c("lightblue", "lightpink")) + ggtitle("NKG7")
g <- FeaturePlot(pbmc, features = "FCER1A", cols = c("lightblue", "lightpink")) + ggtitle("FCER1A")
h <- FeaturePlot(pbmc, features = "FCGR3A", cols = c("lightblue", "lightpink")) + ggtitle("CD16 / FCGR3A")
i <- FeaturePlot(pbmc, features = "S100A8", cols = c("lightblue", "lightpink")) + ggtitle("S100A8")

exp_projection <- d + e + f + g+ h + i + plot_layout(ncol=3)
exp_projection
```

```{r}
top2

```

Here is my assinment of cell type identity to clusters. It's 100% does not reflect reality because the clusters found are poorly defined. But if we compare labeled t-SNE plot and six t-SNE projections of expression markers, we can see that NK cells and T cells are mapped pretty well.

| Cluster ID | Markers           | Cell Type                    |
|------------|-------------------|------------------------------|
| 1          | ANXA2, AQP3       | Epithelial cells             |
| 2          | MAL, CORO1B       | T cells                      |
| 3          | CCR7, LEF1        | Naive T cells                |
| 4          | CD8B, RP11-291B2  | CD8+ T cells                 |
| 5          | FOS, IL7R         | Memory T cells               |
| 6          | GZMK, DUSP2       | Effector memory CD8+ T cells |
| 7          | RP11-290F2, MS4A7 | Monocytes/Macrophages        |
| 8          | FGFBP2, GZMH      | Activated NK cells           |
| 9          | MS4A1, TCL1A      | B cells                      |
| 10         | S100A8, LGALS2    | Neutrophils                  |

```{r}
celltype_mapping <- c(
  "1" = "Epithelial",
  "2" = "T cells",
  "3" = "Naive T",
  "4" = "CD8+ T",
  "5" = "Memory T",
  "6" = "Effector CD8+",
  "7" = "Monocytes/Macrophages",
  "8" = "Activated NK",
  "9" = "B cells",
  "10" = "Neutrophils"
)

pbmc@meta.data$cell_type <- celltype_mapping[as.character(pbmc@meta.data$kmeans_clusters)]

tsne_labeled <- DimPlot(pbmc,
                       reduction = "tsne",
                       group.by = "cell_type",
                       label = TRUE,
                       pt.size = 0.5,
                       repel = TRUE) +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5, face = "bold"))

results_table <- data.frame(
  Cluster = names(celltype_mapping),
  CellType = celltype_mapping,
  Markers = c(
    "ANXA2, AQP3",
    "MAL, CORO1B",
    "CCR7, LEF1",
    "CD8B, RP11-291B2",
    "FOS, IL7R",
    "GZMK, DUSP2",
    "RP11-290F2, MS4A7",
    "FGFBP2, GZMH",
    "MS4A1, TCL1A",
    "S100A8, LGALS2"
  )
)

tsne_labeled
```

```{r}
vln_plot <- vln_plot + 
  theme(
    aspect.ratio = 2, 
    plot.margin = margin(5, 5, 5, 5, "mm")
  ) 

tsne_plot <- tsne_plot + theme(aspect.ratio = 1)
tsne_labeled <- tsne_labeled + theme(aspect.ratio = 1)

exp_projection <- exp_projection + theme(aspect.ratio = 0.2)

combined_plot <- wrap_plots(vln_plot, tsne_plot, tsne_labeled, heatmap, exp_projection,
  nrow = 2,
  widths = c(1, 1, 1),
  heights = c(1, 0.8)
)

ggsave(
  "figure3.png",
  combined_plot,
  width = 25,
  height = 14,
  units = "in",
  dpi = 300
)
```
